% SPM-Kit {\textperiodcentered} Fathom Usage Guide --- v0.1.2
\documentclass[a4paper,11pt]{article}

% == Packages ==
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage[margin=2.2cm, bottom=2.5cm]{geometry}
\usepackage{microtype}
\usepackage[colorlinks=true, linkcolor=blue!60!black, urlcolor=blue!60!black, citecolor=blue!60!black]{hyperref}
\usepackage[table]{xcolor}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{array}
\usepackage{enumitem}
\usepackage[breakable, skins]{tcolorbox}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{upquote}
\usepackage{textcomp}
\usepackage{xurl}
\usepackage{calc}
\usepackage{needspace}
\usepackage{graphicx}
\usepackage{listings}

% == Colours ==
\definecolor{noteblue}{RGB}{30,100,180}
\definecolor{warnamber}{RGB}{180,120,20}
\definecolor{tipgreen}{RGB}{30,130,70}
\definecolor{codegray}{RGB}{50,50,50}
\definecolor{lightgray}{RGB}{245,245,245}

% == Code block style ==
\newcommand{\cmd}[1]{\texttt{\small #1}}

\lstset{
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{lightgray},
  frame=none,
  breaklines=true,
  breakatwhitespace=false,
  postbreak=\mbox{\textcolor{gray}{$\hookrightarrow$}\space},
  showstringspaces=false,
  aboveskip=4pt,
  belowskip=4pt,
  xleftmargin=4pt,
  xrightmargin=4pt,
  framexleftmargin=4pt,
  framexrightmargin=4pt,
  columns=flexible,
  keepspaces=true,
  upquote=true,
}



% == Note / warning boxes ==
\newtcolorbox{infobox}[1][]{colback=noteblue!5, colframe=noteblue!40,
  title={\textcolor{noteblue}{#1}}, fonttitle=\bfseries, breakable}

\newtcolorbox{warnbox}[1][]{colback=warnamber!5, colframe=warnamber!50,
  title={\textcolor{warnamber}{#1}}, fonttitle=\bfseries, breakable}

% == Page style ==
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small SPM-Kit {\textperiodcentered} Fathom --- Usage Guide v0.1.2}
\fancyhead[R]{\small\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0.4pt}

% == Captionless table continuation ==
\newcommand{\tcont}[1]{%
  \multicolumn{1}{c}{\small\dots\ \emph{#1 continued}}\\}

\begin{document}

% ===========================================================
% TITLE
% ===========================================================

\begin{titlepage}
\raggedright
\vspace*{4cm}

{\Huge\bfseries SPM-Kit {\textperiodcentered} Fathom}\\[0.4cm]
{\LARGE Usage Guide}\\[1.5cm]
{\large Version 0.1.2 (Alpha)}\\[0.3cm]
{\large 2026-07-14}\\[2cm]

{\normalsize
  \textbf{Authors:} Jose' Labarca \& Toma's Corrales\\
  \textbf{License:} MIT\\
  \textbf{Repository:} \url{https://github.com/kegouro/spmkit}\\[1cm]

  \emph{Canonical documentation: This PDF is a printable companion. The Markdown
  version at \texttt{docs/user-guide.md} is the authoritative reference and may
  contain additional detail. Both are verified against SPM-Kit~0.1.2.}
}

\vspace{1cm}
{\normalsize
  SPM-Kit is an open-source Python toolkit for analysing AFM/KPFM scanning-probe-microscopy
  data. Fathom is its desktop workspace. This guide covers installation, conceptual model,
  supported formats, the Fathom GUI, CLI reference, Python API, scientific validation,
  and extensibility.
}

\vfill
{\footnotesize
  This open-source project has been developed without dedicated institutional funding.
  Development status: Alpha (3~---~Alpha). API may change before 1.0.
}
\end{titlepage}

\tableofcontents
\clearpage

% ===========================================================
% 1. INTRODUCTION
% ===========================================================
\section{Introduction}

\subsection{What is SPM-Kit?}

SPM-Kit is an open-source Python toolkit for analysing AFM/KPFM scanning-probe-microscopy data.
It reads raw instrument files and produces publication-ready results: roughness
(ISO~25178), KPFM contact-potential statistics, nanomechanical property maps,
thermal-tune cantilever calibration, grain detection, spectral/fractal analysis, and
single-molecule force spectroscopy.

SPM-Kit was developed at the SPM Lab of Universidad Te'cnica Federico Santa Mari'a (UTFSM)
and is released under the MIT license.

\subsection{What is Fathom?}

Fathom is the desktop workspace application bundled with SPM-Kit. It is a PyQt6~+~pyqtgraph
GUI organised into \emph{perspectives} (task-oriented views), not flat tabs. A command
palette (\texttt{Ctrl+K}) provides keyboard access to every function.

\subsection{Component relationship}

\begin{lstlisting}[language=bash]
spmkit (Python package)
  spmkit.core   -- numerical engine (pure Python, no UI imports)
  spmkit.cli    -- command-line interface (Typer + Rich)
  spmkit.gui    -- Fathom desktop workspace (PyQt6 + pyqtgraph)
\end{lstlisting}

\begin{itemize}
  \item \texttt{spmkit} is the Python package and CLI entry point.
  \item \textbf{Fathom} is the graphical workspace.
  \item \textbf{SPM-Kit} is the project/ecosystem name.
\end{itemize}

\subsection{What SPM-Kit does \emph{not} do}
\begin{itemize}[nosep]
  \item It is \textbf{not a microscope controller}.
  \item It is \textbf{not a certified metrology tool}.
  \item It does \textbf{not} provide medical, clinical, or regulatory reports.
  \item Simulation features are \textbf{educational}, not quantitative digital twins.
\end{itemize}

\subsection{Status}
SPM-Kit is \textbf{alpha-stage} (Development Status~3). Core image analysis
(roughness, KPFM, spectral) is validated against Gwyddion at machine precision.
Force-spectroscopy workflows are tested on synthetic and lab data. Some readers
are experimental (see \S\ref{sec:validation}).

% ===========================================================
% 2. INSTALLATION
% ===========================================================
\section{Installation}

Requirements: \textbf{Python\,{\textgreater}=\,3.11}.

\subsection{Core only}
\begin{lstlisting}[language=bash]
pip install spmkit
\end{lstlisting}

Installs: numpy, typer, rich, pyyaml. Sufficient for CLI, Python API, and all image analysis.

\subsection{GUI / Fathom}
\begin{lstlisting}[language=bash]
pip install "spmkit[gui]"
\end{lstlisting}

Adds: PyQt6, pyqtgraph, matplotlib, Crameri colormaps, matplotlib-scalebar.

\subsection{All features}
\begin{lstlisting}[language=bash]
pip install "spmkit[all]"
\end{lstlisting}

\subsection{Extras matrix}

\begin{longtable}{@{}>{\ttfamily}l p{7cm} l@{}}
\toprule
\textnormal{Extra} & \textnormal{Provides} & \textnormal{Deps} \\
\midrule
gui    & Fathom desktop workspace & PyQt6, pyqtgraph, matplotlib \\
viz    & Publication figures, colormaps & matplotlib, cmcrameri, scalebar \\
gwy    & Gwyddion .gwy read/write & gwyfile \\
hdf5   & HDF5 import/export & h5py \\
grains & Grain/particle detection & scipy \\
report & HTML/PDF reports & Jinja2 + spmkit[viz] \\
nanosurf & Validated .nhf reader & NSFopen \\
afm    & JPK QI, .ibw, HDF5, NT-MDT & afmformats \\
jpk    & JPK TIFF force curves & tifffile \\
parallel & Parallel force-volume & joblib \\
pandas & DataFrame export & pandas \\
dev    & Lint, test, type-check & pytest, ruff, mypy, black \\
test-gui & GUI test runner & pytest-qt \\
docs   & Documentation build & mkdocs-material \\
all    & All of the above & --- \\
\bottomrule
\end{longtable}

\subsection{Development installation}
\begin{lstlisting}[language=bash]
git clone https://github.com/kegouro/spmkit
cd spmkit
pip install -e ".[all,dev]"
\end{lstlisting}

\subsection{Headless / HPC}
Install core only: \texttt{pip install spmkit}. Add \texttt{[viz]} for headless figure
generation.

\subsection{Platform notes}
Developed and tested on \textbf{macOS} and \textbf{Linux}. Windows is not regularly tested.
Fathom requires a display; set \texttt{QT\_QPA\_PLATFORM=offscreen} for headless tests.

% ===========================================================
% 3. VERIFICATION
% ===========================================================
\section{Verifying the installation}

\begin{lstlisting}[language=bash]
spmkit --version          # spmkit 0.1.2
python -c "import spmkit; print(spmkit.__version__)"
\end{lstlisting}

% ===========================================================
% 4. CONCEPTUAL MODEL
% ===========================================================
\section{Conceptual model}

\subsection{Architecture}

Three strictly separated layers (enforced by \texttt{tests/test\_architecture.py}):

\begin{itemize}[nosep]
  \item \texttt{core/} --- pure Python, zero UI imports.
  \item \texttt{cli/} --- Typer + Rich. Only imports the public \texttt{core} API.
  \item \texttt{gui/} --- PyQt6 + pyqtgraph. Only imports the public \texttt{core} API.
\end{itemize}

\subsection{Data model}

\begin{lstlisting}[language=bash]
from spmkit import load, SPMData, SPMChannel

data: SPMData = load("scan.nid")
ch: SPMChannel = data["Z-Axis"]

ch.data         # 2D ndarray in physical units
ch.unit         # "m", "V", "{\textdegree}"
ch.x_range      # metres
ch.y_range      # metres
ch.direction    # "forward" | "backward"
\end{lstlisting}

\subsection{Force-spectroscopy domain}

For force curves: \texttt{ForceCurve}, \texttt{ForceSegment}, \texttt{ForceVolume}
(lazy-loaded), \texttt{Calibration}.

\subsection{Result objects}
Every analysis returns an immutable dataclass with a \texttt{.to\_dict()} method,
serialisable via \texttt{spmkit.core.export}.

% ===========================================================
% 5. SUPPORTED FILE FORMATS
% ===========================================================
\section{Supported file formats}

\begin{longtable}{@{}l l l l l@{}}
\toprule
Extension & Source & Type & Access & Status \\
\midrule
.nid & NanoSurf & Images + force-volume & Read & \textbf{Validated} \\
.nhf & NanoSurf HDF5 & Images & Read & Tested \\
.gwy & Gwyddion & Images & Read+Write & Tested \\
.nid (spec) & NanoSurf & Force-volume & Read & Tested \\
.jpk-force & JPK/Bruker & Single curve & Read & Tested \\
.jpk (TIFF) & JPK/Bruker & Force-map/QI & Read & Experimental \\
.001/.002 & Bruker SPM & Images & Read & Experimental \\
.ibw & Asylum & Force curves & Read & Experimental \\
\bottomrule
\end{longtable}

\subsection{Maturity vocabulary}
\begin{description}[nosep]
  \item[Validated] Numerical output cross-checked against independent reference.
  \item[Tested] Unit and synthetic-recovery tests pass.
  \item[Experimental] Reader exists but lacks thorough testing.
\end{description}

\subsection{What ``validated'' means for .nid}
The physical conversion has been verified against Gwyddion exports at
\textbf{correlation~1.000000} for topography, phase, and sensor channels. See
\texttt{docs/TRACEABILITY.md}.

% ===========================================================
% 6. QUICK START
% ===========================================================
\section{Quick start}

\begin{lstlisting}[language=bash]
# Inspect a file
spmkit info scan.nid

# Roughness (ISO 25178)
spmkit roughness scan.nid -c Z-Axis --level plane

# Publication figure
spmkit figure scan.nid -c Z-Axis -o topo.png --colormap batlow

# Launch Fathom
spmkit gui scan.nid

# Force curve fit
spmkit forcecurve curva.jpk-force --model dmt --tip-radius 2e-8
\end{lstlisting}

\needspace{4\baselineskip}
Python:

\begin{lstlisting}[language=bash]
from spmkit import load
from spmkit.core.analysis import leveling, roughness

data = load("scan.nid")
flat = leveling.plane_fit(data["Z-Axis"])
stats = roughness.statistics(flat)
print(f"Sa = {stats.sa:.3g} {stats.unit}")
\end{lstlisting}

% ===========================================================
% 7. FATHOM DESKTOP WORKSPACE
% ===========================================================
\section{Fathom desktop workspace}

\subsection{Launching}
\begin{lstlisting}[language=bash]
spmkit gui               # Fathom (default)
spmkit gui scan.nid      # open file on launch
spmkit gui --legacy      # legacy 7-tab app (fallback)
\end{lstlisting}

\subsection{Opening data}
Drag and drop a file onto the window, or \texttt{Ctrl+O}. Fathom inspects the file and
routes image channels to image perspectives and force/spectroscopy data to force perspectives.
If a file contains both, Fathom asks which to open.

\subsection{Projects (.spmproj)}
\texttt{Ctrl+S} saves the open file, pipeline parameters, and active perspective as a
human-editable YAML file.

\subsection{Appearance}
\begin{itemize}[nosep]
  \item \texttt{Ctrl+Shift+L} --- toggle light/dark theme.
  \item \texttt{Ctrl+Shift+A} --- appearance dialog (theme presets, accent colour, font size).
  \item Settings persist between sessions.
  \item The theme feeds the Qt app, pyqtgraph, and matplotlib simultaneously.
\end{itemize}

\subsection{Command palette}
\texttt{Ctrl+K}, then type a command name and press Enter. Indexes all perspectives and
registered commands.

% ===========================================================
% 8. PERSPECTIVES
% ===========================================================
\section{Perspectives}

Fathom presents 12 perspectives through 6 modules. Each perspective shows only the
panels relevant to its workflow.

\needspace{4\baselineskip}
\subsection{Image perspectives}

\begin{longtable}{@{}>{\ttfamily}l p{9cm}@{}}
\toprule
\multicolumn{2}{c}{\textbf{Module: image}}\\
\midrule
image & Channel viewer, leveling (plane/poly/align-rows), colormap, line profile ROI, roughness + KPFM analysis, profile export.\\
grains & Particle/grain detection on levelled topography; colour overlay, size statistics. Requires \texttt{grains} extra.\\
spectral & Radial PSD (log-log), fractal dimension~D, Hurst exponent~H, correlation length.\\
resonance & Thermal tune: SHO fit $\rightarrow$ f$_0$, Q, spring constant by equipartition.\\
evaporation & Mass sensing: load series of tuning .nid files $\rightarrow$ f(t), mass, d\textsuperscript{2} law fit.\\
\bottomrule
\end{longtable}

\needspace{4\baselineskip}
\subsection{Force perspectives}

\begin{longtable}{@{}>{\ttfamily}l p{9cm}@{}}
\toprule
\multicolumn{2}{c}{\textbf{Module: force}}\\
\midrule
force & Navigate curves, fit Hertz/DMT/Sneddon, inspect modulus $\pm$ 1$\sigma$, adhesion, dissipation, R\textsuperscript{2}. Editable pipeline, Monte Carlo uncertainty. Export scientific CSV.\\
smfs & Single-molecule force spectroscopy: prominence-based rupture detection, WLC/FJC polymer fits, contour-length histogram.\\
map & Force-volume $\rightarrow$ property maps (modulus, adhesion, contact, R\textsuperscript{2}) + histogram. Vectorised (CPU/GPU) or per-curve pipeline.\\
batch & Process folders of force curves $\rightarrow$ summary CSV.\\
\bottomrule
\end{longtable}

\needspace{4\baselineskip}
\subsection{Other perspectives}

\begin{longtable}{@{}>{\ttfamily}l p{9cm}@{}}
\toprule
\multicolumn{2}{c}{\textbf{Module: figure}}\\
\midrule
figure & WYSIWYG publication figure editor: title, axes, colormap, scale bar, draggable annotations. Export PNG/SVG/PDF.\\
\midrule
\multicolumn{2}{c}{\textbf{Module: view3d}}\\
\midrule
view3d & 3D topography surface with hillshade lighting, visual Z exaggeration.\\
\midrule
\multicolumn{2}{c}{\textbf{Module: simulator}}\\
\midrule
simulator & Educational cantilever digital twin: thermal noise spectrum, resonance shift with added mass.\\
\bottomrule
\end{longtable}

% ===========================================================
% 9. END-TO-END WORKFLOWS
% ===========================================================
\section{End-to-end workflows}

\subsection{Topography leveling and roughness}
\textbf{CLI:}
\begin{lstlisting}[language=bash]
spmkit roughness scan.nid -c Z-Axis --level plane
\end{lstlisting}

\textbf{Python:}
\begin{lstlisting}[language=bash]
from spmkit import load
from spmkit.core.analysis import leveling, roughness

data = load("scan.nid")
flat = leveling.plane_fit(data["Z-Axis"])
stats = roughness.statistics(flat)
print(stats.sa, stats.sq, stats.sz, stats.unit)
\end{lstlisting}

\textbf{GUI:} Drag file $\rightarrow$ Image perspective $\rightarrow$ select channel $\rightarrow$ Plane Fit $\rightarrow$ read results from Analysis panel.

\begin{warnbox}[Caveats]
Roughness depends on scan size, pixel density, and leveling method. Always report these
alongside roughness values.
\end{warnbox}

\subsection{Force-curve contact mechanics}

\begin{lstlisting}[language=bash]
spmkit forcecurve sample.jpk-force --model dmt --tip-radius 2e-8
\end{lstlisting}

\textbf{Contact models:} \texttt{sphere} (Hertz), \texttt{paraboloid} (Sneddon paraboloid),
\texttt{cone} (Sneddon cone), \texttt{dmt} (Hertz + adhesion offset). The experimental JKR
model (\texttt{spmkit jkr}) fits adhesive contact but has not been validated against an
independent reference.

\begin{warnbox}[Caveats]
Modulus depends on tip radius (uncertain), contact model (approximation), and calibration.
Report all three alongside the result. JKR is experimental.
\end{warnbox}

\subsection{Force-volume property mapping}

\begin{lstlisting}[language=bash]
spmkit forcemap volume.nid --model dmt --fast -f maps.png
spmkit forceexport volume.nid -o ./results
\end{lstlisting}

Creates property maps (modulus, adhesion, contact point, R\textsuperscript{2}), histograms, and CSV exports.
Use \texttt{--pipeline} for variable-length curves (QI data),
\texttt{--parallel} for multi-core, \texttt{--backend gpu} for GPU (requires CuPy).

\subsection{Thermal tune}
Load a thermal spectrum in the Thermal Tune perspective or use:
\begin{lstlisting}[language=bash]
from spmkit.core.analysis.mechanics import thermal_spring_constant
\end{lstlisting}

\subsection{Evaporation / mass sensing}
\begin{lstlisting}[language=bash]
spmkit evaporation ./tuning/ -k 0.3 -o evap.csv
\end{lstlisting}

Tracks f(t) $\rightarrow$ mass $\rightarrow$ evaporation rate. Fits the d\textsuperscript{2} law for diffusion-limited evaporation.

\subsection{KPFM}
\begin{lstlisting}[language=bash]
spmkit analyze scan.nid --tip-wf 4.8 -o ./results
\end{lstlisting}

Reports mean CPD, standard deviation, and sample work function
($\Phi_{\text{sample}} = \Phi_{\text{tip}} - e\cdot\text{CPD}$).

\subsection{Other workflows}
\begin{lstlisting}[language=bash]
spmkit grains scan.nid --min-size 10           # grain detection
spmkit psd scan.nid -c Z-Axis                   # fractal analysis
spmkit batch ./measurements/ -o summary.csv     # batch processing
spmkit fbatch ./curves/ -o batch.csv            # force batch
spmkit figure scan.nid -o topo.png --colormap batlow  # figure export
\end{lstlisting}

% ===========================================================
% 10. COMMAND-LINE REFERENCE
% ===========================================================
\section{Command-line reference}

18 commands are registered in the Typer application.

\begin{longtable}{@{}>{\ttfamily}l p{8.5cm}@{}}
\toprule
Command & Description \\
\midrule
info & Show file metadata and channels \\
roughness & ISO 25178 parameters (Sa, Sq, Sz, Ssk, Sku) \\
psd & Spectral: fractal dimension, Hurst, correlation length \\
analyze & Full pipeline: roughness + KPFM $\rightarrow$ CSV + JSON \\
nanomech & Force-curve fit (Hertz/DMT/Sneddon) $\rightarrow$ Young's modulus \\
grains & Grain/particle detection and size statistics \\
batch & Process folder $\rightarrow$ summary CSV \\
evaporation & Mass sensing: f(t) $\rightarrow$ mass, evaporation rate, d\textsuperscript{2} law \\
figure & Publication figure: PNG/SVG/PDF with scale bar \\
convert & Format conversion (.nid $\rightarrow$ .gwy / .h5) \\
verify & .nid byte-level integrity audit (8 checks) \\
gui & Launch Fathom (or legacy with --legacy) \\
workspace & Force-curve workspace (redesign) \\
forcecurve & Fit single force curve (JPK/NanoSurf) \\
forcemap & Force-volume $\rightarrow$ property maps \\
forcereport & Force-volume report (HTML/LaTeX/PDF) \\
forceexport & Full export bundle to folder \\
fbatch & Batch force processing $\rightarrow$ CSV \\
jkr & \textbf{[Experimental]} JKR adhesive contact fit \\
\bottomrule
\end{longtable}

\subsection{Common options}
\begin{itemize}[nosep]
  \item \texttt{-{}-channel / -c} --- channel name (default: \texttt{Z-Axis} for images,
    \texttt{Deflection} for force)
  \item \texttt{-{}-output / -o} --- output path
  \item \texttt{-{}-level / -l} --- leveling: \texttt{plane}, \texttt{poly}, \texttt{none}
  \item \texttt{-{}-model} --- contact model: \texttt{sphere}, \texttt{paraboloid}, \texttt{cone}, \texttt{dmt}
  \item \texttt{-{}-tip-radius} --- tip radius in metres (default: \texttt{1e-08})
\end{itemize}

Run \texttt{spmkit <command> --help} for full options.

% ===========================================================
% 11. PYTHON API
% ===========================================================
\section{Python API}

\subsection{Public imports}

\begin{lstlisting}[language=bash]
from spmkit import load, SPMData, SPMChannel, __version__
from spmkit.core.analysis import leveling, roughness, kpfm, mechanics
from spmkit.core.analysis import spectral, grains, resonance
from spmkit.core.export import to_csv, to_json, to_hdf5
from spmkit.core.viz import FigureSpec, save_figure
\end{lstlisting}

\subsection{Loading and inspecting}

\begin{lstlisting}[language=bash]
data = load("scan.nid")    # auto-detect format
print(data.names)          # channel names
ch = data["Z-Axis"]        # SPMChannel
\end{lstlisting}

\subsection{Analysis}

\begin{lstlisting}[language=bash]
flat = leveling.plane_fit(data["Z-Axis"])
stats = roughness.statistics(flat)
cpd = kpfm.statistics(data["CPD"], tip_work_function=4.8)
frac = spectral.fractal_dimension(flat)
grains_result = grains.detect(flat, min_size=4)
\end{lstlisting}

\subsection{Force spectroscopy}

\begin{lstlisting}[language=bash]
from spmkit.core.io.forceload import load_nid_force
from spmkit.core.analysis.mechanics import fit_hertz

volume = load_nid_force("sample.nid")
result = fit_hertz(volume.curves[0], tip_radius=10e-9, model="dmt")
print(result.young_modulus, result.r_squared)
\end{lstlisting}

\subsection{Export and figures}

\begin{lstlisting}[language=bash]
to_csv(stats, "roughness.csv")
to_json(stats, "roughness.json")
to_hdf5(data, "scan.h5")

spec = FigureSpec(title="AFM Topography", colormap="batlow")
save_figure(flat, spec, "topo.png")
\end{lstlisting}

\subsection{Recipes}

\begin{lstlisting}[language=bash]
from spmkit.core.pipeline import Recipe, Step, run

recipe = Recipe(steps=(
    Step(op="calibrate"),
    Step(op="find_contact_point"),
    Step(op="fit_elasticity", params={"model": "dmt", "tip_radius": 2e-8}),
))
_, ctx = run(recipe, curve)
print(ctx["young_modulus"], ctx["r_squared"])
\end{lstlisting}

% ===========================================================
% 12. KEYBOARD SHORTCUTS
% ===========================================================
\section{Keyboard shortcuts}

\begin{tabularx}{\linewidth}{@{}>{\ttfamily}l X@{}}
\toprule
Shortcut & Action \\
\midrule
Ctrl+K & Command palette \\
Ctrl+O & Open file \\
Ctrl+S & Save project \\
Ctrl+M & Calculate map \\
Ctrl+Shift+R & Generate report \\
Ctrl+Shift+L & Toggle light/dark theme \\
Ctrl+Shift+A & Customize appearance \\
Ctrl+Left / Ctrl+Right & Previous / next curve \\
Ctrl+1 ... Ctrl+5 & Tab switch (legacy app only) \\
\bottomrule
\end{tabularx}

% ===========================================================
% 13. EXPORTS AND PROVENANCE
% ===========================================================
\section{Exports and provenance}

\subsection{Export formats}

\begin{tabularx}{\linewidth}{@{}>{\ttfamily}l l X@{}}
\toprule
Format & Via & Notes \\
\midrule
CSV & \texttt{to\_csv()}, CLI & Units in headers; empty cells for NaN \\
JSON & \texttt{to\_json()}, CLI & Structured; programmatic consumption \\
HDF5 & \texttt{to\_hdf5()}, convert & Self-describing binary; requires hdf5 extra \\
.gwy & \texttt{spmkit convert} & Round-trip with Gwyddion \\
PNG/SVG/PDF & \texttt{spmkit figure} & Publication-quality figures \\
HTML/PDF & \texttt{spmkit forcereport} & Full force-volume report \\
YAML & \texttt{core.pipeline} & Reproducible pipeline recipe \\
\bottomrule
\end{tabularx}

\subsection{Provenance}
\begin{itemize}[nosep]
  \item \texttt{SPMData.source\_path} records the original file path.
  \item \texttt{SPMData.metadata} preserves instrument parameters.
  \item \texttt{spmkit verify} performs 8 byte-level integrity checks on .nid files.
  \item YAML recipes capture the exact pipeline parameters.
  \item Exports include metadata when the format supports it.
\end{itemize}

% ===========================================================
% 14. SCIENTIFIC VALIDATION
% ===========================================================
\section{Scientific validation}
\label{sec:validation}

SPM-Kit uses a layered approach:

\begin{enumerate}[nosep]
  \item \textbf{Unit tests} --- function correctness on synthetic input.
  \item \textbf{Synthetic recovery} --- known-property data $\rightarrow$ verify recovery.
  \item \textbf{Cross-checks} --- compare against Gwyddion for the same input.
  \item \textbf{Instrument-file loading} --- real lab files load without errors.
  \item \textbf{Validation} --- numerical agreement at defined tolerance (only .nid image channels).
  \item \textbf{Experimental} --- exists but not validated or extensively tested.
\end{enumerate}

\subsection{Status summary}

\begin{longtable}{@{}p{3.3cm} p{2.8cm} p{1.8cm} p{3.5cm}@{}}
\toprule
Capability & Evidence & Status & Limitations \\
\midrule
.nid image parsing & vs Gwyddion: corr 1.000000 & \textbf{Validated} & Signed 32-bit LE; other dtypes untested \\
.nid physical conv. & 8 integrity checks & \textbf{Validated} & Assumes Dim2Range linear \\
.nid orientation & flipud for Y channels & \textbf{Validated} & Non-image channels as-is \\
Roughness ISO 25178 & Synthetic $\sigma$ recovery & Tested & Plane-levelled data only \\
KPFM CPD statistics & Unit tests & Tested & Work function: tip calibration \\
Hertz/DMT force fit & Synthetic recovery & Tested & Tip radius uncertainty dominant \\
Force-volume maps & Unit + synthetic tests & Tested & Grid regularity assumed \\
Thermal tune & Unit tests & Tested & Equipartition assumption \\
Evaporation & Unit tests & Tested & Spherical droplet; diffusion-limited \\
Grain detection & Unit tests & Tested & Threshold-based; no shape metrics \\
Fractal/PSD & Unit tests & Tested & Self-affine model; finite bandwidth \\
JKR adhesive contact & No external ref & \textbf{Experimental} & Use with caution \\
.nhf reader & Unit tests & Tested & Generic HDF5 walker \\
.gwy reader/writer & Round-trip test & Tested & Verified \\
.jpk-force reader & Unit tests & Tested & Calibration from metadata \\
JPK QI (.jpk TIFF) & Limited testing & \textbf{Experimental} & Requires jpk extra \\
Bruker SPM (.001) & Limited testing & \textbf{Experimental} & Requires afm extra \\
afmformats readers & Limited testing & \textbf{Experimental} & Quality varies by format \\
SMFS polymer fits & Unit tests & Tested & Model choice critical \\
Fathom GUI & 37 GUI tests & Tested & Qt version/platform dependent \\
Architecture & AST test & Enforced & No runtime enforcement \\
\bottomrule
\end{longtable}

\subsection{Gwyddion as reference}
Gwyddion serves as an \emph{independent reference implementation}, not universal ground
truth. Both tools implement their own conversion from the same byte stream.

\subsection{Simulator}
The cantilever simulator is \textbf{educational only}. It models a damped harmonic
oscillator with additive noise. Not suitable for calibration or metrology.

% ===========================================================
% 15. SAFETY AND INTERPRETATION
% ===========================================================
\section{Safety and interpretation warnings}

\begin{itemize}
  \item \textbf{Results require domain judgment.} High R\textsuperscript{2} does not guarantee correct model choice.
  \item \textbf{Calibration values matter.} Young's modulus depends on tip radius (uncertain),
    spring constant (calibration), and contact model (approximation).
  \item \textbf{Check units.} \texttt{SPMChannel.unit} tells the physical unit.
  \item \textbf{Experimental readers may misinterpret metadata.} Format parsing is
    best-effort for reverse-engineered formats.
  \item \textbf{Fitted parameters are not automatically physical truth.}
  \item \textbf{SPM-Kit is not a medical, clinical, regulatory, or safety-critical instrument.}
  \item \textbf{Preserve original data.} SPM-Kit reads without modifying files.
\end{itemize}

% ===========================================================
% 16. EXTENSIBILITY
% ===========================================================
\section{Extensibility}

\subsection{Entry-point groups}

\begin{tabularx}{\linewidth}{@{}>{\ttfamily}l X@{}}
\toprule
Group & Purpose \\
\midrule
spmkit.plugins.v1 & Readers, analyses, domains \\
spmkit.gui.modules & Fathom modules (perspectives + panels) \\
\bottomrule
\end{tabularx}

\subsection{Adding a format}
Implement the \texttt{Reader} Protocol (\texttt{inspect + load}) and register via
\texttt{spmkit.plugins.v1}.

\subsection{Adding a perspective}
Define a \texttt{ModuleSpec} with \texttt{PanelSpec} and \texttt{PerspectiveSpec},
register via \texttt{spmkit.gui.modules}. See \texttt{docs/extending.md} for full details.

% ===========================================================
% 17. TROUBLESHOOTING
% ===========================================================
\section{Troubleshooting}

\begin{description}[nosep, font=\normalfont\ttfamily]
  \item[Missing GUI extra] \hfill \\ \texttt{pip install "spmkit[gui]"}
  \item[Qt platform plugin error] \hfill \\ Install Qt dependencies or set \texttt{QT\_QPA\_PLATFORM=offscreen}.
  \item[Unsupported format] \hfill \\ Check extension matches \texttt{.nid/.nhf/.gwy/.jpk-force}. For experimental readers, install \texttt{afm} or \texttt{jpk}.
  \item[Missing optional dependency] \hfill \\ Install the relevant extra (e.g. \texttt{pip install "spmkit[grains]"}).
  \item[Large force-volume memory] \hfill \\ Use lazy loading or the \texttt{--fast} CPU vectorised path.
  \item[GPU fallback] \hfill \\ CuPy is not bundled. Install separately: \texttt{pip install cupy-cuda12x}.
  \item[Export failure] \hfill \\ Check write permissions. Use \texttt{--output} with a writable path.
\end{description}

% ===========================================================
% 18. DEVELOPMENT
% ===========================================================
\section{Development and quality}

\begin{lstlisting}[language=bash]
git clone https://github.com/kegouro/spmkit
cd spmkit
pip install -e ".[all,dev,test-gui]"
pre-commit install
\end{lstlisting}

\textbf{Quality checks (CI gate):}
\begin{lstlisting}[language=bash]
make check        # lint + types + tests
make lint         # ruff check
make type         # mypy (strict on core)
make test         # pytest with coverage
\end{lstlisting}

\textbf{GUI tests} (headless):
\begin{lstlisting}[language=bash]
QT_QPA_PLATFORM=offscreen pytest tests/gui -q
\end{lstlisting}

\textbf{Documentation:}
\begin{lstlisting}[language=bash]
pip install -e ".[docs]"
python -m mkdocs build --strict
\end{lstlisting}

% ===========================================================
% 19. DATA HUNTER
% ===========================================================
\section{Relationship to SPM-Kit Data Hunter}

SPM-Kit Data Hunter is a companion project that discovers and curates public data
candidates for validation. It does \textbf{not} analyse data and does \textbf{not}
validate SPM-Kit output.

\begin{itemize}[nosep]
  \item \textbf{SPM-Kit / Fathom} --- analyse data.
  \item \textbf{Data Hunter} --- discover data for validation.
  \item \textbf{Discovery is not validation.}
\end{itemize}

% ===========================================================
% 20. CITATION AND LICENSE
% ===========================================================
\section{Citation, license, and contributions}

\subsection{How to cite}

\begin{lstlisting}[language=bash]
@software{spmkit2026,
  author       = {Jose' Labarca and Toma's Corrales},
  title        = {spmkit: Open-source AFM/KPFM analyser
                  for scanning probe microscopy},
  year         = {2026},
  version      = {0.1.2},
  url          = {https://github.com/kegouro/spmkit},
}
\end{lstlisting}

\subsection{License}
MIT. See \texttt{LICENSE} in the repository root.

\subsection{Contributing}
Issues and pull requests at \url{https://github.com/kegouro/spmkit/issues}.
See \texttt{CONTRIBUTING.md}.

\subsection{Funding}
This open-source project has been developed without dedicated institutional funding.

\subsection{Project status}
\textbf{Alpha (0.1.2).} Core functionality is stable and tested. APIs may change before
1.0. New contributors and format readers are welcome.

% ===========================================================
% APPENDICES
% ===========================================================
\appendix

\section{Quick reference card}

\begin{lstlisting}[language=bash]
# Install
pip install "spmkit[gui]"

# Inspect
spmkit info scan.nid

# Analyse
spmkit roughness scan.nid -c Z-Axis --level plane
spmkit psd scan.nid -c Z-Axis
spmkit analyze scan.nid -o ./results --tip-wf 4.8

# Figures
spmkit figure scan.nid -o topo.png --colormap batlow

# Convert
spmkit convert scan.nid scan.gwy

# Verify
spmkit verify scan.nid

# Force
spmkit forcecurve curve.jpk-force --model dmt --tip-radius 2e-8
spmkit forcemap volume.nid --fast -f maps.png
spmkit forcereport volume.nid -o report
spmkit forceexport volume.nid -o ./export
spmkit fbatch ./curves/ -o batch.csv

# Evaporation
spmkit evaporation ./tuning/ -k 0.3 -o evap.csv

# Grains
spmkit grains scan.nid --min-size 10

# GUI
spmkit gui [file]
\end{lstlisting}

\section{CLI command index}

\begin{longtable}{@{}>{\ttfamily}l l l@{}}
\toprule
Command & Category & Input \\
\midrule
info & Inspection & .nid, .nhf \\
roughness & Image & .nid, .nhf, .gwy \\
psd & Spectral & .nid, .nhf, .gwy \\
analyze & Pipeline & .nid, .nhf, .gwy \\
nanomech & Force & .nid (spectroscopy) \\
grains & Image & .nid, .nhf, .gwy \\
batch & Batch & Folder of SPM \\
evaporation & Resonance & Folder of .nid tuning \\
figure & Figure & .nid, .nhf, .gwy \\
convert & Format & .nid \\
verify & Audit & .nid \\
gui & GUI & --- (optional file) \\
workspace & GUI & Force-curve file \\
forcecurve & Force & .jpk-force, .nid \\
forcemap & Force & .nid (force-volume) \\
forcereport & Force & .nid (force-volume) \\
forceexport & Force & .nid (force-volume) \\
fbatch & Batch & Folder of force curves \\
jkr & Force (exp.) & Calibrated curve \\
\bottomrule
\end{longtable}

\section{Perspective index}

\begin{longtable}{@{}>{\ttfamily}l l l l@{}}
\toprule
Key & Label & Module & Type \\
\midrule
image & Imagen & image & Image \\
grains & Granos & image & Image \\
spectral & Espectral & image & Image \\
resonance & Sintoni'a te'rmica & image & Image \\
evaporation & Evaporacio'n & image & Image \\
force & Curva de fuerza & force & Force \\
smfs & SMFS & force & Force \\
map & Mapa & force & Force \\
batch & Batch & force & Force \\
figure & Figura & figure & Figure \\
view3d & Vista 3D & view3d & Image \\
simulator & Simulador & simulator & Educational \\
\bottomrule
\end{longtable}

\section{Supported formats matrix}

\begin{longtable}{@{}l l l l l@{}}
\toprule
Extension & Source & Type & Access & Status \\
\midrule
.nid & NanoSurf & Images + FV & Read & Validated \\
.nhf & NanoSurf HDF5 & Images & Read & Tested \\
.gwy & Gwyddion & Images & Read+Write & Tested \\
.nid (spec) & NanoSurf & Force-volume & Read & Tested \\
.jpk-force & JPK/Bruker & Single curve & Read & Tested \\
.jpk (TIFF) & JPK/Bruker & Force-map/QI & Read & Experimental \\
.001/.002 & Bruker SPM & Images & Read & Experimental \\
.ibw & Asylum & Force curves & Read & Experimental \\
\bottomrule
\end{longtable}

\section{Keyboard shortcuts}

\begin{tabularx}{\linewidth}{@{}>{\ttfamily}l X@{}}
\toprule
Shortcut & Action \\
\midrule
Ctrl+K & Command palette \\
Ctrl+O & Open file \\
Ctrl+S & Save project \\
Ctrl+M & Calculate map \\
Ctrl+Shift+R & Generate report \\
Ctrl+Shift+L & Toggle theme \\
Ctrl+Shift+A & Appearance settings \\
Ctrl+Left / Ctrl+Right & Previous / next curve \\
Ctrl+1 ... Ctrl+5 & Tab switch (legacy app) \\
\bottomrule
\end{tabularx}

\vfill

\begin{center}
{\small --- End of SPM-Kit {\textperiodcentered} Fathom Usage Guide v0.1.2 ---}
\end{center}

\end{document}
